---
title: 'Part 4: Characterization of F4 on genomic and epigenomic level'
author: "Junyan Lu"
date: "`r doc_date()`"
output:
  BiocStyle::html_document
---

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone")}

plotDir = ifelse(exists(".standalone"), "", "part4/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
knitr::opts_chunk$set(fig.path=plotDir, dev=c("png", "pdf"))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

#Libraries
library(mofaCLL)
library(gridExtra)
library(ggrepel)
library(ggbeeswarm)
library(GEOquery)
library(DESeq2)
library(MOFA)
library(grImport)
library(fgsea)
library(limma)
library(MultiAssayExperiment)
library(pheatmap)
library(cowplot)
library(tidyverse)
```


# Association between F4 (CLL-PD) and genomics

Load datasets
```{r}
data("mofaOut","gene","rna","mutLoad","tabFACS")
```

Download methylation dataset from server to a temporary folder
```{r, eval=!exists(".standalone")}
methPath <- file.path(tempdir(),"meth.RData")
if(!file.exists(methPath)) {
  download.file("https://www.huber.embl.de/users/jlu/data/meth.RData",
                methPath)
  load(methPath)
} else {
  load(methPath)
}
```



Process MOFA model
```{r}
#get factor values
facTab <- getFactors(
  MOFAobject, 
  factors = "LF4",
  as.data.frame = TRUE
) %>% as_tibble() 

#get all factors
facTab.all <- getFactors(
  MOFAobject,
  factors = "all",
  as.data.frame = TRUE
) %>% as_tibble()

#get weights
weightTab <- getWeights(
  MOFAobject,
  factors = "all",
  as.data.frame = TRUE
) %>% as_tibble() %>%
  mutate(feature = ifelse(feature == "IGHV.status", "IGHV", feature))
#get training data
trainData <- getTrainData(MOFAobject)

detach("package:MOFA", unload = TRUE)
```


## Loading of features in the genomic view
```{r,fig.width=6, fig.height=6}
plotGeneWeight <- plotTopWeights.m(weightTab, view =  "Mutations", factor = "LF4", nfeatures = 5) + ylab("Absolute loading on F4 (CLL-PD)") +
  theme(plot.margin = margin(10,0,10,0))
plotGeneWeight
```

## Test of associations between genetic variations with latent factor

Student's t-test
```{r}
sampleOverlap <- intersect(rownames(gene.unfiltered), facTab$sample)
genData <- gene.unfiltered[sampleOverlap,]

#remove gene with higher than 40% missing values
genData <- genData[,colSums(is.na(genData))/nrow(genData) <= 0.4]

#remove genes with less than 3 mutated cases
genData <- genData[,colSums(genData, na.rm = TRUE) >= 3]

#t-test
y <- facTab[match(sampleOverlap, facTab$sample),]$value
tRes <- apply(genData, 2, function(x) {
  res <- t.test(y ~ as.factor(x), var.equal = TRUE)
  data.frame(p = res$p.value, 
             df = res$estimate[[2]] - res$estimate[[1]])
}) %>% bind_rows() %>% mutate(gene = colnames(genData),
                              p.adj = p.adjust(p, method = "BH")) %>%
  arrange(p)
filter(tRes, p.adj <0.05) %>% mutate_if(is.numeric, formatNum, digits =3, format = "e") %>% DT::datatable()
```

Volcano plot
```{r, fig.height=5, fig.width=6}
plotGeneVolcano <- plotVolcano(tRes, posCol = colList[1], negCol = colList[2],
            x_lab = "Difference of mean", ifLabel = TRUE) + 
  theme(legend.position = "none",
        plot.margin = margin(8,8,8,8),
        axis.title.y = element_text(vjust=0))
plotGeneVolcano
```

Boxplots of associated genes (5% FDR)
```{r, fig.height=10, fig.width=10}
tRes.sig <- filter(tRes, p.adj <= 0.05)
plotList <- lapply(tRes.sig$gene, function(geneName) {
   plotTab <- tibble(factor = y, 
                     status = genData[,geneName]) %>%
     filter(!is.na(status)) %>%
     mutate(status = ifelse(status ==1, "mut","wt")) %>%
     group_by(status) %>% mutate(n=n()) %>% ungroup() %>%
     mutate(status = sprintf("%s\n(N=%s)",status,n)) %>%
     arrange(desc(n)) %>% mutate(status = factor(status, levels = unique(status))) %>%
     mutate(geneName = italicizeGene(geneName))
   pval <- formatNum(filter(tRes, gene == geneName)$p, digits = 1, format="e")
   titleText <- sprintf("%s~' ('~italic(P)~'='~'%s'~')'",unique(plotTab$geneName),pval) 
   ggplot(plotTab, aes(x=status, y = factor)) +
     geom_boxplot(width=0.3, aes(fill = status), outlier.shape = NA) + 
     geom_beeswarm(col = "black", size =1,cex = 1.5, alpha=0.5) +
     ggtitle(parse(text = titleText))+
     #ggtitle(sprintf("%s (p = %s)",geneName, formatNum(pval, digits = 1, format = "e"))) + 
     ylab("F4 value") + xlab("") +
     scale_fill_manual(values = colList[3:4]) +
     theme_full +
     theme(legend.position = "none",
           plot.title = element_text(hjust = 0.5),
           plot.margin = margin(3,3,3,3))
})
grid.arrange(grobs = plotList, ncol=3)
plotGeneViolin <- plotList
```

Heatmap plot showing associated genetic features (5% FDR), arranged by F4 values.   
```{r,fig.width=8, fig.height=4}
annoCol <- data.frame(row.names = sampleOverlap,
                      F4 = facTab[match(sampleOverlap, facTab$sample),]$value)
annoCol <- annoCol[order(annoCol$F4),,drop = FALSE]
plotMat <- t(genData[rownames(annoCol),filter(tRes, p.adj <= 0.05)$gene])
plotMat[is.na(plotMat)] <- 0
breaks <- c(0,0.5,1)
colors <- c("white","black")
rowLabs <- italicizeGene(rownames(plotMat))
plotGeneHeatmap <- pheatmap(plotMat, cluster_cols = FALSE, cluster_rows = FALSE, 
                            annotation_col = annoCol, border_color = "black",
         breaks = breaks, color = colors, show_colnames = FALSE, 
         labels_row = parse(text = rowLabs),
         legend = FALSE, silent = TRUE)$gtable
plot_grid(plotGeneHeatmap)
```

## Correlation with the mutation load

## Scatter plot of correlation with mutations load and recurrent mutation load

```{r}
compareTab <- filter(mutLoadSum, group == "Overall", subgroup == "Overall") %>%
  select(patID, platform, n)
sumReCurr.cnv <- gene.unfiltered[,c("trisomy12","del13q","del11q","del17p",
                                    "gain2p","gain8q","del8p","del6q","del18p","del20p")]
sumRecurr.snv <- gene.unfiltered[,71:113] #only select recurrent gene mutations
sumRecurr <- cbind(sumRecurr.snv, sumReCurr.cnv) %>%
  data.frame() %>% rownames_to_column("patID") %>%
  gather(key = "gene", value = "status",-patID) %>%
  group_by(patID) %>% summarise(n.recurr = sum(status,na.rm = TRUE))

compareTab <- left_join(compareTab, sumRecurr, by = "patID") %>% 
  mutate(LF4 = facTab[match(patID, facTab$sample),]$value) %>%
  filter(!is.na(LF4),!is.na(n),!is.na(n.recurr))
```

## Assocations with total number of mutations

Function to plot and annotate assocations
```{r}
plotCor <- function(plotTab, x, y, x_label, y_label, title, color = "black") {
  corRes <- cor.test(plotTab[[x]], plotTab[[y]], use = "pairwise.complete.obs")
  annoCoef <- paste("'coefficient ='~",format(corRes$estimate,digits = 2))
  annoP <- paste("italic(P)~'='~",formatNum(corRes$p.value, digits = 1, format = "e"))
  
  ggplot(plotTab, aes_string(x = x, y = y)) + 
    geom_point(shape = 21, fill =color, size=3) + 
    geom_smooth(method = "lm", se=FALSE, color = "grey50", linetype ="dashed" ) + 
    annotate("text", x = max(plotTab[[x]]), y = Inf, label = annoCoef,
           hjust=1, vjust =1.5, size = 5, parse = TRUE, col= colList[1]) +
    annotate("text", x = max(plotTab[[x]]), y = Inf, label = annoP,
           hjust=1, vjust =3, size = 5, parse = TRUE, col= colList[1]) +
    ylab(y_label) + xlab(x_label) + ggtitle(title) +
    theme_full + theme(plot.margin = margin(8,8,8,8))
}
```

#### WES
```{r, fig.width=5, fig.height=5}
plotTab <- filter(compareTab, platform == "WES") 
corTotal.wes <- plotCor(plotTab, "LF4", "n", "CLL-PD", "Total number of mutations",
        sprintf("WES dataset (n=%s)",nrow(plotTab)), colList[[6]])
corTotal.wes
```

#### WGS
```{r, fig.width=5, fig.height=5}
plotTab <- filter(compareTab, platform == "WGS") 
corTotal.wgs <- plotCor(plotTab, "LF4", "n", "CLL-PD", "Total number of mutations",
        sprintf("WGS dataset (n=%s)",nrow(plotTab)), colList[[6]])
corTotal.wgs

```

### Assocations with recurrent aberration

```{r, fig.width=5, fig.height=5}
plotTab <- distinct(compareTab, patID, .keep_all = TRUE)
corRecurr <- plotCor(plotTab, "LF4", "n.recurr", "CLL-PD", "Number of recurrent aberrations",
        "",colList[[5]]) 
corRecurr
```


# Assocation between F4 (CLL-PD) and DNA methylation

## Correlation test using linear model

```{r}
methMat <- assays(meth[,colnames(meth) %in% facTab$sample])[["beta"]]
factorMatrix <- facTab.all %>% spread(key = factor, value = value) %>%
  data.frame() %>% column_to_rownames("sample") 
factorMatrix <- factorMatrix[colnames(methMat),]
factorMatrix <- factorMatrix[,complete.cases(t(factorMatrix))]
designMat <- model.matrix(~1 + .,factorMatrix)
fit <- lmFit(methMat, designMat)
fit2 <- eBayes(fit)
corRes <- limma::topTable(fit2, number ="all", coef = "LF4", adjust.method = "BH") %>%
  rownames_to_column("id") %>% as_tibble()
corRes.LF1 <- limma::topTable(fit2, number ="all", coef = "LF1", adjust.method = "BH") %>%
  rownames_to_column("id") %>% as_tibble()
```

Number of tested associations
```{r}
dim(methMat)
```

## Compare number of significant correlations in F1 and F4

```{r, fig.height=4, fig.width=6}
numTab.LF4 <- filter(corRes, adj.P.Val <= 0.01) %>% mutate(direction = ifelse(t>0, "pos","neg")) %>%
  group_by(direction) %>% summarise(number = length(id)) %>%
  mutate(factor = "F4 (CLL-PD)")
numTab.LF1 <- filter(corRes.LF1, adj.P.Val <= 0.01) %>% mutate(direction = ifelse(t>0, "pos","neg")) %>%
  group_by(direction) %>% summarise(number = length(id)) %>% 
  mutate(factor = "F1")

plotTab <- bind_rows(numTab.LF4, numTab.LF1)
sigNumPlot <- ggplot(plotTab, aes(x=factor, y=number, fill = direction)) + 
  geom_bar(stat = "identity", width = 0.8,
           position = position_dodge2(width = 6),
           col = "black") +
  geom_text(aes(label=number), 
            position = position_dodge(width = 0.9),
            size=4, hjust=-0.1)  +
  scale_fill_manual(name = "", labels = c("Hypomethylated","Hypermethylated"), values = colList) +
  coord_flip(ylim = c(0,70000), expand = FALSE) + xlab("") + ylab("Number of significant associations") + theme_half +
  theme(legend.position = c(0.75,0.22), legend.text = element_text(size=12), legend.background = element_rect(fill = NA))
sigNumPlot
```


## Correlation between mean methylation values and F4
```{r, fig.width=5, fig.height=5}
meanBeta <- colMeans(methMat)
plotTab <- tibble(patID = names(meanBeta),
                  meanBeta = meanBeta,
                  LF4 = facTab[match(patID, facTab$sample),]$value)

corBeta <- cor.test(plotTab$meanBeta, plotTab$LF4)

annoCoef <- paste("'coefficient ='~",format(corBeta$estimate,digits = 2))
annoP <- paste("italic(P)~'='~",formatNum(corBeta$p.value, digits = 1, format = "e"))

plotMeanBeta <- ggplot(plotTab, aes(x = LF4, y = meanBeta)) + 
  geom_point(shape = 21, fill =colList[3], size=3) + 
  geom_smooth(method = "lm", se=FALSE, color = "grey50", linetype ="dashed" ) + 
  annotate("text", x = 3, y = Inf, label = annoCoef,
           hjust=1, vjust =1.5, size = 5, parse = TRUE, col= colList[1]) +
  annotate("text", x = 3, y = Inf, label = annoP,
           hjust=1, vjust =3, size = 5, parse = TRUE, col= colList[1]) +
  ylab("Mean beta values") + xlab("CLL-PD") +
  theme_full 

plotMeanBeta
```

## Enrichment of F4 associated probes in the PMD region

### Probes in the CLL specific PMD region

Read list of CGs that lie in the PMD region, according to Mallm et al. 2019 (PMID: 31118277) 
```{r}
probesPMD <- readLines(system.file("externalData/PMD_probes.txt", package = "mofaCLL"))
```

How many CpGs correlated with LF4 are in the PMD region?
```{r}
sigProbe <- filter(corRes, adj.P.Val <=0.01, t <0)$id
table(sigProbe %in% probesPMD)
```

Fisher's exact test
```{r}
testTab <- table(rownames(meth) %in% sigProbe, rownames(meth) %in% probesPMD)
fisher.test(testTab, alternative = "greater")
```

### Probes in the general solo-WCGW PMD region

Download the list of CG probes in the solo-WCGW PMD region (Zhou et al., 2018, PMID: 29610480)
```{r}
wcgwPath <- system.file("externalData/hm450.comPMD.probes.tsv", package = "mofaCLL")
probesPMD.wcgw <- read_delim(wcgwPath, delim = "\t", col_names = FALSE, skip = 1)[[2]]
```

How many probes annotated as WCGW probes are on our array?
```{r}
onArray <- probesPMD.wcgw %in% rownames(meth)
table(onArray)
#subset
probesPMD.wcgw <- probesPMD.wcgw[onArray]
```

How many CpGs correlated with LF4 are in the WCGW region?
```{r}
table(sigProbe %in% probesPMD.wcgw)
```

Fisher's exact test
```{r}
testTab <- table(rownames(meth) %in% sigProbe, rownames(meth) %in% probesPMD.wcgw)
fisher.test(testTab, alternative = "greater")
```

FGSEA tset and plot
```{r WCGWenrich, fig.height=4, fig.width=6}
sigList <- list("Solo-WCGW CpGs in PMDs" = probesPMD.wcgw)
enrichRes <- runFGSEA(corRes, sigList, pcut =1, maxSize = 10000, name = "id", stat = "t")
enrichPlot <- enrichRes$plots[[1]] + theme(plot.margin = margin(1,20,1,20))
enrichPlot
```


## Association between CLL-PD(F4) and CLL proliferation ability (Ki-67 populaiton)

### Paired t-test

```{r}
testTab <- tabFACS %>% mutate(posKI = CD19posKI67pos)
waterTab <- filter(testTab, treat == "H2O") %>% select(patID, posKI) %>%
  dplyr::rename(popWater = posKI)
testTab <- filter(testTab, treat!="H2O") %>%
  dplyr::rename(popTreat = posKI) %>% left_join(waterTab)
```

```{r}
resPair <- group_by(testTab, treat, groupLF) %>% nest() %>%
  mutate(m = map(data,~t.test(.$popTreat,.$popWater, paired = TRUE))) %>%
  mutate(res = map(m, broom::tidy)) %>% unnest(res) %>%
  select(treat, groupLF, p.value) %>% arrange(p.value)
resPair
```


### T-test for compare CLL-PD high and low for CpG treatment

#### Univariate test
```{r}
testTab <- tabFACS %>% filter(treat == "CPG") %>% mutate(posKI = CD19posKI67pos)
resUni <- t.test(posKI ~ groupLF, testTab)
resUni
```

#### Multi-vairate test
```{r}
resMulti <- car::Anova(lm(posKI ~ groupLF + IGHV, testTab))
resMulti
```

### Box plots to show associations

```{r, fig.height=5, fig.width=6}
plotTab <- tabFACS %>% mutate(posKI = CD19posKI67pos/(CD19posKI67pos+CD19posKI67neg)) %>%
  mutate(subtype = plyr::revalue(IGHV, c("M"="M-CLL","U"="U-CLL")),
         group = plyr::revalue(groupLF, c("high" = "high CLL-PD\n(n=12)","low"="low CLL-PD\n(n=12)")),
         treat = plyr::revalue(treat, c("H2O" = "water", "CPG" = "CpG ODN"))) %>%
  mutate(group = factor(group,levels = c("high CLL-PD\n(n=12)","low CLL-PD\n(n=12)")),
         treat = factor(treat, levels = c("water","CpG ODN"))) %>%
  arrange(posKI) %>% mutate(patID = factor(patID, levels = unique(patID)))

annoText <- tibble(tt = c(paste("italic(P)~'=",formatNum(resPair$p.value[1], digits = 2),"'"),
                          paste("italic(P)~'=",formatNum(resPair$p.value[2], digits = 2),"'")), 
                   group  = c("high CLL-PD\n(n=12)","low CLL-PD\n(n=12)"))

plotCpG <- ggplot(plotTab, aes(x=treat, y = CD19posKI67pos)) + #geom_boxplot(outlier.alpha = NA) + 
  geom_point(aes(fill = IGHV), shape =21, size =4, cex=4, alpha=0.8) +
  geom_segment(x =1, xend =2, y=40,yend=40, size=0.2) +
  geom_line(aes(x=treat,y=CD19posKI67pos, group = patID), col = "grey50", linetype = "dashed") +
  geom_text(data =annoText, aes(label = tt), x = 1.5, y =40, vjust=-1, parse=TRUE, size=5) +
  scale_fill_manual(values = colList[c(3,6)]) +
  facet_wrap(~group) +
  ylim(c(0,55)) + theme_half+
  theme(legend.position = "bottom", strip.text = element_text(size=15, face= "bold"),
        strip.background = element_rect(fill = "white",color = "white"),
        axis.text.x = element_text(size=15.5),
        legend.text = element_text(size=15),
        legend.title = element_text(size=13),
        legend.margin = margin(0,0,0,0),
        legend.box.margin=margin(-5,-5,-5,-5)) +
  ylab("% Ki-67+ CD19+ cells") + xlab("")
plotCpG
```

## Organizing plots for Section 3 in the manuscript

### Main Figure 3
```{r figure3, fig.height=10, fig.width=16}
enMyc <- pictureGrob(readPicture(system.file("externalData//MYC.eps.xml", package = "mofaCLL")))

title = ggdraw() + draw_figure_label("Figure 3", fontface = "bold", position = "top.left",size=22)

pout <- plot_grid(plot_grid(plotGeneWeight, plotGeneVolcano, ncol=1, labels = c("a","b"), 
                            rel_heights = c(0.4,0.6), align = "hv", axis = "lr", label_size = 22),
                  plot_grid(corRecurr + theme(axis.title.y = element_text(vjust =0)),
                             plotMeanBeta , ncol=1, labels = c("c","e"), rel_heights = c(5,5),align = "hv", axis = "lr", label_size = 22),
          plot_grid(NULL, sigNumPlot,enMyc, plotCpG, ncol=1, labels = c("d","","f","g"), rel_heights = c(0.05,0.3,0.2,0.45), 
                    align = "hv", label_size = 22),
          ncol=3,rel_widths = c(1,0.9,1))

plot_grid(title, pout, rel_heights = c(0.05,0.95), ncol = 1)
```

### Supplementary figures

#### Associations between genomic variantions and LF4
```{r genomics_LF, fig.height=16, fig.width=12}
gridViolin <- plot_grid(plotlist = plotGeneViolin, ncol =3)
pout <- ggdraw() +
  draw_plot(plotGeneHeatmap, 0, 0.70, 1, 0.30) +
  draw_plot(gridViolin, 0, 0 , 1, 0.69)
pout
```


#### Associations between LF4 and mutation load
```{r mutLoad_LF, fig.height=4.5, fig.width=9}
plot_grid(corTotal.wes, corTotal.wgs,
          labels = c(" "," "), label_size = 22)
```

