---
title: 'Part 6: Single cell analysis using CyTOF'
output:
  BiocStyle::html_document
---

# Single cell analysis (CyTOF)

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone")}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

plotDir = ifelse(exists(".standalone"), "", "part6/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
knitr::opts_chunk$set(fig.path=plotDir, dev=c("png", "pdf"))

#load packages and dataset
library(CATALYST)
library(diffcyt)
library(cowplot)
library(mofaCLL)
library(glmnet)
library(RColorBrewer)
library(ComplexHeatmap)
library(scattermore)
library(tidyverse)
```

Load cyTOF dataset
```{r}
rm(list = ls()) #clean environment

cytoPath <- file.path(tempdir(),"sceAll.RData")
if(!file.exists(cytoPath)) {
  download.file("https://www.huber.embl.de/users/jlu/data/sceAll.RData",
                cytoPath)
  load(cytoPath)
} else {
  load(cytoPath)
}

md <- metadata(sceAll)$experiment_info %>% as_tibble()
```


## T-SNE plots

### T-SNE of all samples and cells, colored by cell types
```{r TSNE_all, fig.height=7.3, fig.width=7.3}
reTab <- getReducedDimTab(sceAll,"TSNE")
plotTab <- reTab %>% 
  mutate(cluster = factor(clustMap[as.character(stateCluster)], clustMap))
p<-plotDM(plotTab, sceAll,colorBy = "cluster", x = "x",y="y", plotTitle = FALSE) +
  guides(color = guide_legend(ncol= 2,override.aes = list(alpha = 1, size = 6))) +
  theme(legend.text = element_text(size=28), legend.title = element_text(size=20))
p
```

### Proliferation markers, stratified by CLL-PD

#### Ki-67 (for main figure)
```{r Ki_67_CLLPD, fig.height=4, fig.width=8}
useDim <- "TSNE"
sceCpG <- sceAll[,!is.na(reducedDim(sceAll,useDim)[,1])]
sceCpG <- sceCpG[,sceCpG$condition == "CpG"]
dmTab  <- getReducedDimTab(sceCpG,useDim)

p <- plotDM(dmTab, sceCpG, "Ki-67", facetBy = "CLL-PD", facetCol = 2, x = "x", y =  "y", fast=FALSE, pointsize =3.2) +
  theme(legend.text = element_text(size=22), legend.title = element_text(size=22))
p
```

#### P-Rb & Cyclin B1
```{r pRb_cyclinB1_CLLPD, fig.height=8, fig.width=8}
pList <- list()

pList[[1]] <- plotDM(dmTab, sceCpG, "P-Rb", facetBy = "CLL-PD", facetCol = 2, x = "x", y =  "y",  fast=FALSE, pointsize =3.2)+
  theme(legend.text = element_text(size=20), legend.title = element_text(size=20))
pList[[2]] <- plotDM(dmTab, sceCpG, "Cyclin B1", facetBy = "CLL-PD", facetCol = 2, x = "x", y =  "y", fast=FALSE, pointsize =3.2)

p <- shareLegend(pList, ncol=1, position = "bottom")

p
```


### Pooled samples, colored by markers associated with proliferation, under CpG condition
```{r driver6_expr, fig.height=5, fig.width=12}
useDim <- "TSNE"
sceCpG <- sceAll[,!is.na(reducedDim(sceAll,useDim)[,1])]
sceCpG <- sceCpG[,sceCpG$condition == "CpG"]
dmTab  <- getReducedDimTab(sceCpG,useDim)
markerNames <- c("CDK4","GLUT1","c-Myc","P-4E-BP1","P-AMPK alpha","NFAT1")

plotList <- lapply(markerNames, function(n) {
   plotDM(dmTab, sceCpG, n, x = "x", y =  "y", fast=FALSE, pointsize = 4) +
   theme(legend.text = element_text(size=20), 
         legend.title = element_text(size=20),
         plot.margin = margin(0,1.8,0,1.8,"cm"))
})

p<-shareLegend(plotList,position = "bottom", ncol=3)
p
```

#### Separate for CLL-PD high and low samples
```{r driver6_expr_CLLPD, fig.height=15, fig.width=18}
plotList <- lapply(markerNames, function(n) {
   plotDM(dmTab, sceCpG, n, x = "x", y =  "y", fast=FALSE, facetBy = "CLL-PD", facetCol = 2, pointsize = 4) +
   theme(legend.text = element_text(size=20), legend.title = element_text(size=20),
         plot.margin = margin(1,1,1,1, unit = "cm"))
})

p <- shareLegend(plotList,position = "bottom", ncol=2)
p
```


## Differential population abundance

### Test for differential abundance
```{r}
sceViable <- sceAll[,sceAll$typeCluster == "CLL"]
popTab <- getPop(sceViable,"stateCluster") %>% left_join(md, by = "sample_id")
rm(sceViable)
```

```{r, warning=FALSE, message=FALSE}
condiList <- levels(sceAll$condition)

resDA_CLLPD <-lapply(condiList, function(eachCondi) {
  #subset
  sceSub <- subsetSCE(sceAll, eachCondi)
  
  #define experiment design
  ei <- metadata(sceSub)$experiment_info
  ei$CLLPD <- relevel(ei$CLLPD, ref = "low")
  designMat <- createDesignMatrix(ei, cols_design = c("CLLPD"))
  contrast <- createContrast(c(0, 1))

  #test using diffcyte
  ds_res <- diffcyt(sceSub, 
    design = designMat, contrast = contrast, 
    analysis_type = "DA", method_DA = "diffcyt-DA-edgeR",
    clustering_to_use = "stateCluster", verbose = TRUE,
    transform = FALSE)
  
  #get results
  res <- rowData(ds_res$res) %>%
    data.frame() %>% mutate(condition = eachCondi)
  res
}) %>% bind_rows()
```

```{r}
plotPopCLLPD <- lapply(condiList, function(eachCondi) {
  eachList <- lapply(unique(popTab$cluster), function(clusterName) {
    eachTab <- popTab %>% dplyr::filter(condition %in% eachCondi, cluster == clusterName) %>%
      mutate(condition = as.character(condition)) %>%
      mutate(condiName = condiMap[eachCondi],
             CLLPD = factor(paste0(CLLPD," CLL-PD"), levels = c("high CLL-PD","low CLL-PD")))
    pVal <- dplyr::filter(resDA_CLLPD, cluster_id == clusterName, condition == eachCondi)$p_val
    
    annoP <- data.frame(annoP = ifelse(pVal < 1e-12, paste("italic(P)~'<",formatNum(1e-12, digits = 1),"'"),
                    paste("italic(P)~'=",formatNum(pVal, digits = 1),"'")))
    
    yMax <- max(eachTab$popSize)
    
    eachPlot <- ggplot(eachTab, aes(x= CLLPD, y = popSize)) + 
              #geom_boxplot(width=0.5, outlier.shape = NA) +
              ggbeeswarm::geom_quasirandom(aes(fill = IGHV), shape =21, cex=5, width = 0.3) +
              ylim(c(0, yMax + yMax*0.5)) +
              geom_segment(x =1, xend = 2, y=yMax + yMax*0.1 ,yend= yMax + yMax*0.1, size=0.8) +
              scale_fill_manual(values = colorMap, name = "IGHV status") + 
              geom_text(data = annoP, aes(label = annoP),x = 1.5, y= yMax + yMax*0.1, vjust=-1, parse=TRUE, size=6) +
              ylab("% in live CLL cells") + xlab("") + ggtitle(clusterName) +
              theme_half  +
              theme(legend.position = "bottom",
                    axis.text = element_text(size=15),
                    axis.title = element_text(size=15),
                    legend.box.margin=margin(-20,0,0,0),
                    legend.margin=margin(0,0,0,0),
                    legend.text = element_text(size=15),
                    legend.title = element_text(size=15)) 
    })
  names(eachList) <- unique(popTab$cluster)
  eachList
})
names(plotPopCLLPD) <- condiList
```

### Proliferating fraction under CpG condition
```{r pop_CLLPD, fig.height=5, fig.width=4}
p <- plotPopCLLPD$CpG$CLL_proliferating
pPopCLLPD <- p+ggtitle("proliferating CLL cells\n(CLL-PD high vs low)")  + ylim(0,50)

#pPopCLLPD

```

### Proliferating fraction change under all four conditons

```{r}
plotPopCondition <- function(popTab, condiList, clusterList, colorBy = "CLL-PD",
                             yLabel = "% in all cells", plotTitle = "", yLim  = NULL) {
  
  plotTab <- popTab %>% dplyr::filter(condition %in% condiList, cluster %in% clusterList) %>%
    mutate(condition = as.character(condition)) %>% 
    mutate(condiName = factor(condiMapS[condition],levels = condiMapS[condiList])) %>%
    mutate(clusterName = factor(clustMap[as.character(cluster)], levels = clustMap[clusterList]))
  
  eachPlot <- ggplot(plotTab, aes(x= condiName, y = popSize))
      
  if (colorBy == "CLL-PD") {
    eachPlot <- eachPlot + 
      geom_line(aes(group = patient_id, col = CLLPD), linetype = "solid", size=1) +
      geom_point(aes(col = CLLPD), size = 3) 
  } else {
    eachPlot <- eachPlot + geom_point(aes(col = IGHV), size=3)  +
            geom_line(aes(group = patient_id, col = IGHV), linetype = "dashed", size=1) 
  }
      
  eachPlot <- eachPlot + 
      #scale_fill_manual(values = colorMap, name = colorBy) + 
      scale_color_manual(values = colorMap, name = colorBy) + 
      expand_limits(y=c(0,yLim)) +
      ylab(yLabel) + xlab("")
  
  if (length(clusterList) > 1) {
    eachPlot <- eachPlot + facet_wrap(~clusterName, ncol=2) 
  } else {
    eachPlot <- eachPlot + ggtitle(plotTitle)
  }
  eachPlot <- eachPlot  + theme_half  +
      theme(legend.position = "bottom", 
            strip.text = element_text(size=15, face= "bold"),
            strip.background = element_rect(fill = "white",color = "white"),
            axis.text.x = element_text(size=14),
            legend.box.margin=margin(-20,0,0,0),
            legend.margin=margin(0,0,0,0),
            legend.text = element_text(size=15),
            legend.title = element_text(size=15)) 
  

  
  return(eachPlot)
}
```

```{r clusterSize_Pro_viable, fig.width=8, fig.height=5}
clusterList <- c("CLL_proliferating")
condiList <- c("DMSO","CpG","CpG_Ever","Ever")
pPopCondi <- plotPopCondition(popTab, condiList, clusterList, yLabel = "% in live CLL cells", plotTitle = "proliferating CLL cells\n(across conditions)") + ylim(0,50)
#pPopCondi
```

Combine and align axis for figure c and d
```{r proliferation_pop, fig.height=4, fig.width=9}
plot_grid(pPopCondi, 
          pPopCLLPD + theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(),
                            axis.title.y = element_blank()),
          align = "hv", axis = "rb", rel_widths = c(0.65,0.35))
```


## Differential marker expression

Subset for CLL population
```{r}
sceCLL <- subsetSCE(sceAll, cluster_id = "CLL", k = "typeCluster")
#rm(sceAll)
#select functional marker (non-linage and non-proliferating)
markerUse <- setdiff(state_markers(sceCLL),c("Ki-67","Cyclin B1","P-Rb"))
sceCLL <- sceCLL[rownames(sceCLL) %in% markerUse, ]
```


### Heatmap plot to summarise changes

```{r}
getMedian <- function(sceObj, useCluster) {
  clustTab <- metadata(sceObj)$cluster_code
  allCluster <- clustTab[match(sceObj$cluster_id, clustTab$som100),][[useCluster]]
  exprMat <- assays(sceObj)[["exprs"]]

  sumTab <- lapply(seq(nrow(exprMat)), function(i) {
    data.frame(expr = exprMat[i,],
               sample_id = sceObj$sample_id,
               cluster = allCluster) %>%
      group_by(sample_id, cluster) %>%
      summarise(medVal = median(expr, na.rm=TRUE),
                .groups = "drop") %>%
      mutate(antigen = rownames(exprMat)[i])
  }) %>% bind_rows()
  return(sumTab)
}
```

```{r}
exprTab <- getMedian(sceCLL, useCluster = "typeCluster") %>% left_join(md) %>%
  mutate(exprVal = medVal)
sumTab <- group_by(exprTab, condition, CLLPD, antigen) %>%
  summarise(val = median(exprVal)) %>%
  mutate(colID = paste0(condition, "_",CLLPD)) %>%
  ungroup()
```


```{r}
colAnno <- distinct(sumTab, condition, CLLPD, colID) %>%
  mutate(condition = factor(condition, levels = c("DMSO","CpG","CpG_Ever","Ever")),
         CLLPD = factor(CLLPD, levels = c("high","low"))) %>%
  arrange(condition, CLLPD) %>%
  mutate(Condition = condiMapS[as.character(condition)],
         `CLL-PD` = CLLPD) %>%
  select(colID, Condition, `CLL-PD`) %>%
  tibble() %>% column_to_rownames("colID")

exprMat <- select(sumTab, colID, antigen, val) %>%
  pivot_wider(names_from = colID, values_from = val) %>%
  data.frame() %>% column_to_rownames("antigen")
exprMat <- exprMat[,rownames(colAnno)]
```

Prepare annotation rows
```{r}
annoColor <- list(`CLL-PD` = c(high = colList[3], low=colList[6]))
annoTab <- colAnno %>% as_tibble(rownames = "colID") %>%
  mutate(Condition = factor(Condition, levels = unique(Condition)))

pCondi <- ggplot(annoTab, aes(x=Condition, y= "  Condition")) +
  geom_tile(fill = "white",color = "black", size=0.5, alpha=0) +
  scale_y_discrete(position = "right", expand = c(0,0)) +
  coord_cartesian(expand = FALSE) +
  geom_text(aes(label = Condition),size=4.5, fontface = "plain", lineheight =0.8) + theme_void() +
  theme(axis.text.y = element_text(size=12, face = "bold"),
        panel.background = element_rect(size=1, color = "black"))

pPd <- ggplot(annoTab, aes(x=colID, y= "  CLL-PD")) +
  geom_tile(aes(fill = `CLL-PD`),color = "black", size=0.5, alpha=0.5) +
  scale_y_discrete(position = "right", expand = c(0,0)) +
  coord_cartesian(expand=FALSE) +
  scale_fill_manual(values = annoColor$`CLL-PD`) +
  geom_text(aes(label = `CLL-PD`),size=4.5, fontface = "plain") + theme_void() +
  theme(axis.text.y = element_text(size=12, face = "bold"), legend.position = "none",
        panel.background = element_rect(size=1, color = "black"),
        plot.margin = margin(1,0,2,0))


pColAnno <- plot_grid(pCondi, pPd, ncol=1, align = "hv",axis = "lr", rel_heights = c(1.4,1))
pColAnno
```


```{r expr_heatmap, fig.height=5.5, fig.width=8}
exprMat <- mscale(exprMat, center = TRUE, scale = FALSE, useMad = TRUE)
pHeat <- pheatmap::pheatmap(as.matrix(exprMat), scale = "none", cluster_cols = FALSE, 
                   gaps_col = c(2,4,6,8), border_color = "grey30",
                   treeheight_row = 20, clustering_method = "ward.D2", fontsize = 12, fontsize_row = 15,
                   show_colnames = FALSE, breaks = seq(-0.8,0.8, length.out = 100),
                   color = colorRampPalette(c(colList[2],"white",colList[1]))(101), silent = TRUE)$gtable
plot_grid(plot_grid(NULL, pColAnno,  NULL, rel_widths = c(0.47, 8, 2.1), nrow=1),
          pHeat, nrow=2, rel_heights = c(0.15,1))
```

### Differential marker expression related to condition

#### In all CLL cells

Define all the necessary comparison groups and clusters to use
```{r}
#comparison list
compareList <- list(c("DMSO","CpG"),
                 c("CpG", "CpG_Ever"))

#cluster to use
clusterUse <- "typeCluster"
```


#### Performing test
```{r}
resDE_condition <- diffCondiDE(sceCLL, compareList, clusterUse) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))
```

#### Volcano plots

```{r}
plotVolcanoList <- function(resDE_condition, compareList, fdrCut = 0.05, clusterPlot = NULL, useFdr = TRUE,
                            globalPcut = NULL) {

  if (is.null(clusterPlot)) clusterPlot <- unique(resDE_condition$cluster_id)

  plotList <- lapply(clusterPlot, function(eachCluster) {

    eachList <- lapply(compareList, function(eachPair) {
      plotTab <- dplyr::filter(resDE_condition, cluster_id == eachCluster, control == eachPair[1], treat == eachPair[2])
      p <- plotCyToVolcano(plotTab, fdrCut = fdrCut, labSize = 5.2,
                           plotTitle = sprintf("%s vs %s",condiMap[eachPair[2]], condiMap[eachPair[1]]), 
                           useFdr=useFdr, globalPcut = globalPcut)
      p
    })

  })
  names(plotList) <- clusterPlot
  return(plotList)
}

```


##### DMSO, CpG, CpG + Ever
```{r volcano_plot, fig.width=11, fig.height=6}
fdrCut = 0.05
globalPcut <- max(filter(resDE_condition, p_adj <= fdrCut)$p_val)
volList <- plotVolcanoList(resDE_condition, compareList, globalPcut = globalPcut)

shareLegend(list(volList$CLL[[1]] + ylim(0, 9) + xlim(-1.7,1.7),
                 volList$CLL[[2]] + ylim(0, 9) + xlim(-1.7,1.7) + 
                   theme(axis.text.y = element_blank(),
                         axis.ticks.y = element_blank(),
                         axis.title.y = element_blank())), 
            position = "bottom", ratio = c(1,0.05),
            rel_widths = c(0.55,0.5))
```



```{r}
plotExprConditionFacet <- function(exprTab, markers, conditions, resDE_condition, clusterPlot = "CLL",ncol=3) {
    
  #prepare table for plot
  plotTab <- dplyr::filter(exprTab, cluster == clusterPlot, antigen %in% markers, condition %in% conditions) %>%
    mutate(antigen = factor(antigen, levels = markers),
           condiName = factor(condiMapS[as.character(condition)],levels = condiMapS[conditions]))
  
  pTab <- lapply(seq(length(conditions) -1), function(i) {
    tab <- dplyr::filter(resDE_condition, marker_id %in% markers, control == conditions[i], treat == conditions[i+1]) %>%
      mutate(x1 = i+0.1, x2 = i+0.9, xp = i+0.5)
    return(tab)
  }) %>% bind_rows() %>%
    select(marker_id, p_val, control, treat, x1, x2, xp)
  
  rangeTab <- group_by(plotTab, antigen) %>% summarise(yMax = max(medVal), yMin = min(medVal)) %>%
    mutate(yRange = yMax-yMin)
  
  pTab <- left_join(pTab, rangeTab, by = c(marker_id="antigen")) %>%
          mutate(annoP = ifelse(p_val < 1e-12, paste("italic(P)*'<",formatNum(1e-12, digits = 1),"'"),
                                paste("italic(P)*'=",formatNum(p_val, digits = 1),"'"))) %>%
    mutate(antigen = marker_id)
  
  scaleFUN <- function(x) sprintf("%.1f", x)

  p<-ggplot(plotTab, aes(x=condiName, y=exprVal)) + 
    geom_point(aes(col = CLLPD), size=3) +
    geom_line(aes(group = patient_id, col = CLLPD), linetype = "solid",size=1) +
    scale_y_continuous(labels=scaleFUN, limits = c(0,NA), expand = expand_scale(add = c(0.1,0.5)))+
    geom_blank(data = pTab, aes(x=1, y=yMax + yRange*0.5)) +
    geom_segment(data = pTab, aes(x=x1, xend =x2, y = yMax+yRange*0.3, yend = yMax+yRange*0.3)) +
    geom_text(data = pTab, aes(label = annoP, x=xp, y=yMax+yRange*0.3), vjust=-1, parse=TRUE, size=4) +
    guides(col = guide_legend(override.aes = list(alpha = 1, size = 3))) +
    scale_color_manual(values = colorMap, name = "CLL-PD") + 
    ylab("Median intensity") + xlab("")+
    facet_wrap(~antigen, scale = "free_y") +
    theme_half  + theme(legend.position = "bottom", 
                        strip.text = element_text(size=15, face= "bold"),
                        strip.background = element_rect(fill = "white",color = "white"),
                        axis.text.x = element_text(size=12),
                        legend.text = element_text(size=13),
                        legend.title = element_text(size=13),
                        legend.box.margin=margin(-20,0,0,0),
                        legend.margin=margin(0,0,0,0))
  
  return(p)
}
```

#### Key markers 
```{r plot3Cond, fig.height = 4.5, fig.width = 10}
plotExprConditionFacet(exprTab, c("CDK4","GLUT1","c-Myc","P-4E-BP1","P-S6K","P-AMPK alpha"),c("DMSO","CpG","CpG_Ever"), resDE_condition)
```

## Differential marker expression related to CLL-PD

### Using all cell type populations

```{r}
#cluster to use
clusterUse <- "typeCluster"

#cluster to plot 
clusterPlot <- c("CLL")

#sce object for testing
sceTest <- subsetSCE(sceCLL, condiList = c("CpG","DMSO"))

exprTab <- filter(exprTab, condition %in% c("CpG","DMSO"))
```


#### Performing test
```{r}
resDiffPd <- diffPdDE(sceTest,clusterUse) %>% mutate(p_adj = p.adjust(p_val, method = "BH"))
```

#### Summary of results

### Visualize selected samples
```{r diffBoxExpr,fig.height=4, fig.width=10}
markerList <- c("CDK4","GLUT1","c-Myc", "P-4E-BP1","P-S6K","P-AMPK alpha")
p <- plotDiffExpr(exprTab, markerList, resDiffPd, useCondi = "CpG")
p
```


#### Volcano plots (Supplementary)

Volcano plot
```{r}
plotVolcanoListPD <- function(resDE_CLLPD, fdrCut = 0.05, clusterPlot = NULL, useFdr = FALSE) {
  
  if (is.null(clusterPlot)) clusterPlot <- unique(resDE_CLLPD$cluster_id)
  
  plotList <- lapply(clusterPlot, function(eachCluster) {
    
    eachList <- lapply(unique(resDE_CLLPD$condition), function(eachCondi) {
      plotTab <- dplyr::filter(resDE_CLLPD, cluster_id == eachCluster, condition == eachCondi)
      p <- plotCyToVolcano(plotTab, fdrCut = fdrCut, 
                       plotTitle = sprintf("high vs low CLL-PD (%s)", condiMap[eachCondi]),
                       useFdr = useFdr)
      p
    })
    names(eachList) <- unique(resDE_CLLPD$condition)
    return(eachList)
  })
  names(plotList) <- clusterPlot
  return(plotList)
}
pVolPD <- plotVolcanoListPD(resDiffPd)
```

#### Upon CpG treatment

```{r suppl_volcano_CLLPD_CPG, fig.height = 6, fig.width = 6}
resDiffPd.cpg <- filter(resDiffPd, condition == "CpG") %>% 
  mutate(p_adj = p.adjust(p_val, method = "BH"))

pVolPD <- plotVolcanoListPD(resDiffPd.cpg, fdrCut = 0.1, useFdr = TRUE)

p1 <- pVolPD$CLL$CpG + ylim(0, 3) + xlim(-0.8,0.8)

p1
```

```{r}
rm(sceTest)
rm(sceCLL)
```

## Identify markers (other than proliferating markers) that differentiate proliferatign and non-proliferating cells


```{r, echo=TRUE}
cellSample <- function(sceObj, useCluster, nCell, poolSample = FALSE) {
  
  sceObj$index <- seq(ncol(sceObj))
  
  if (!poolSample) {
    sampleTab <- colData(sceObj) %>% data.frame() %>%
      group_by(sample_id, !!sym(useCluster)) %>% sample_n(nCell) 
  } else{
    sampleTab <- colData(sceObj) %>% data.frame() %>%
      group_by(!!sym(useCluster)) %>% sample_n(nCell) 
  }
  
  sceObj <- sceObj[,sceObj$index %in% sampleTab$index]
  sceObj <- sceObj[, sample(seq(ncol(sceObj)))]
  
  return(sceObj)
}

#Functions for running glm
identifyMarker <- function(sceObj, useCluster, useMarker, nCell = 500, poolSample=FALSE, 
                            repeats=20, folds = 10, lambda = "lambda.1se") {
  
  sceObj <- sceObj[rownames(sceObj) %in% useMarker,]
  
  modelList <- list()
  lambdaList <- c()
  alpha = 1
  coefMat <- matrix(NA, nrow(sceObj), repeats)
  rownames(coefMat) <- rownames(sceObj)
  
  for (i in seq(repeats)) {
      
    #sample cells
    sceSub <- cellSample(sceObj, useCluster,nCell,poolSample)

    
    y <- droplevels(sceSub[[useCluster]])
    X <- t(assay(sceSub))
    
    res <- cv.glmnet(X,y, type.measure = "auc", family="binomial", 
                     nfolds = folds, alpha = alpha, standardize = TRUE)
    
    lambdaList <- c(lambdaList, res[[lambda]])
    
    modelList[[i]] <- res
      
    coefModel <- coef(res, s = lambda)[-1] #remove intercept row
    coefMat[,i] <- coefModel
  }
  list(modelList = modelList, lambdaList = lambdaList, coefMat = coefMat)
}
```

```{r}
#Function for the heatmap plot
lassoPlotMarker <- function(lassoOut, sceObj, annoColor, freqCut = 0, coefCut = 0.01, useCluster = "stateCluster", 
                            annoMarker = c("Ki-67","Cyclin B1","P-Rb")) {
  
  
  #for the barplot on the left of the heatmap
  barValue <- rowMeans(lassoOut$coefMat)
  freqSeq <- rowMeans(lassoOut$coefMat !=0)
  
  #filter
  barValue <- barValue[barValue > coefCut & freqSeq >= freqCut]
  
  barValue <- barValue[order(barValue, decreasing = TRUE)]

  #for the heatmap and scatter plot below the heatmap
  colnames(sceObj) <- paste0("c",seq(ncol(sceObj)))
  exprMat <- t(scaleExprs(t(assay(sceObj))))
  #exprMat <- mscale(exprMat, censor =5)
  
  #prepare column annotation table
  
  colAnnoTab <- data.frame(row.names = colnames(exprMat),
                           state = sceObj[[useCluster]])
  
  annoMarkerTab <- exprMat[annoMarker,]
  colAnnoTab <- cbind(colAnnoTab, t(annoMarkerTab))
  colAnnoTab <- colAnnoTab[order(colAnnoTab$state),]
  
  #remove annotation markers
  exprMat <- exprMat[!rownames(exprMat) %in% annoMarker, ]
  exprMat <- exprMat[names(barValue),] #reorder row by coefficient
  exprMat <- exprMat[,rownames(colAnnoTab)] #order column by cluster
  #construct heatmap
  
  
  
  myCol <- colorRampPalette(c('blue','yellow','red'), 
                 space = "Lab")
  
  haCol <- ComplexHeatmap::HeatmapAnnotation(df = colAnnoTab,  col = annoColor,
                                             which = "column",
                                             show_legend = c(FALSE,TRUE,TRUE,TRUE),
                                             annotation_name_gp = gpar(fontface = "bold"),
                                             annotation_legend_param = list(
                                               grid_height = unit(0.6,"cm"),
                                               title_gp = gpar(cex=1.1)
                                             ), simple_anno_size = unit(0.7,"cm"))
  
  #bar annotation on the left
  annoBar = rowAnnotation(`importance score` = anno_barplot(barValue, baseline = 0, border = FALSE,width = unit(3.5,"cm"),
                                               gp = gpar(fill = "grey80"),
                                               axis_param = list(direction = "reverse"),
                                               at = c(0,0.5,1)))
  
  ComplexHeatmap::Heatmap(exprMat, col = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100),
                          name = "scaled intensity", top_annotation = haCol,
                          left_annotation = annoBar, show_column_names = FALSE,
                          #top_annotation = haCol, left_annotation = haRow, show_column_names = FALSE,
                          cluster_columns  = FALSE, cluster_rows = FALSE,
                          column_title = "",
                          heatmap_legend_param = list(title_position =  "lefttop-rot", grid_height = unit(0.8,"cm"),
                                                      title_gp = gpar(cex=1.1)),
                          use_raster = TRUE)
                        
}
```

### In CLL-PD high samples with proliferation

Sample equal number of cells from proliferating and non-proliferating compartment

```{r}
cMap <- c(CLL_resting = "non-proliferating",CLL_proliferating = "proliferating")
sceSub <- sceAll[,sceAll$typeCluster =="CLL" & sceAll$condition == "CpG" & sceAll$CLLPD == "high" & sceAll$patient_id!="H272"]
sceSub$stateCluster <- factor(cMap[as.character(sceSub$stateCluster)], levels = cMap)

useMarker <- setdiff(state_markers(sceSub),c("Ki-67","Cyclin B1","P-Rb"))
rm(sceAll)
```

```{r}
set.seed(1118)
lassoOut <- identifyMarker(sceSub, "stateCluster", useMarker, nCell = 500, repeats = 100, poolSample = FALSE)
```


```{r}
sceVis <- cellSample(sceSub, "stateCluster", 500)
```

```{r lasso_CLLPD_high, fig.height=4.5, fig.width=10}
col_fun1 = circlize::colorRamp2(c(0,1), c("white",colList[4]))
col_fun2 = circlize::colorRamp2(c(0,1), c("white",colList[6]))
col_fun3 = circlize::colorRamp2(c(0,1), c("white",colList[5]))
annoColor <- list(state = c(proliferating = "#BD3D2A4E", `non-proliferating` = "#0071B87E"), 
                  `Ki-67` = col_fun3,
                  `Cyclin B1` = col_fun3,
                  `P-Rb` = col_fun3)

lassoPlotMarker(lassoOut, sceVis, annoColor = annoColor, coefCut = 0, freqCut = 0.8)
```
